---
title: "Bike_trip_Chicago_2019"
output: ipynb_notebook
---


# Problem Statement
In this case study, I use Divvy trips data set which has data of a bike-sharing company in Chicago during 2019. Analyzing the given data set we aim to design marketing strategies aimed at converting casual riders into annual members.

Data collected from [Divvy_trips](https://divvy-tripdata.s3.amazonaws.com/index.html)


# Importing necessary libraries
```{r warning=FALSE, , echo=TRUE}
library(tidyverse)
library(lubridate)
library(ggplot2)
```

# Reading datasets of Bike trips in 2019 over 4 quarters
```{r}
df1=read_csv("D:/Capstone dataset/Divvy_Trips_2019_Q1.csv")
df2=read_csv("D:/Capstone dataset/Divvy_Trips_2019_Q2.csv")
df3=read_csv("D:/Capstone dataset/Divvy_Trips_2019_Q3.csv")
df4=read_csv("D:/Capstone dataset/Divvy_Trips_2019_Q4.csv")
```

# Making data consistent across all 4 datasets
```{r echo=TRUE}
colnames(df1)
colnames(df2)
colnames(df3)
colnames(df4)
df2 <- df2 %>% 
  rename(
    trip_id = "01 - Rental Details Rental ID"
          ,bikeid = "01 - Rental Details Bike ID" 
          ,start_time = "01 - Rental Details Local Start Time"  
          ,end_time = "01 - Rental Details Local End Time"  
          ,tripduration=  "01 - Rental Details Duration In Seconds Uncapped"
          ,from_station_name = "03 - Rental Start Station Name" 
          ,from_station_id = "03 - Rental Start Station ID"
          ,to_station_name = "02 - Rental End Station Name" 
          ,to_station_id = "02 - Rental End Station ID"
          ,usertype = "User Type"
          ,gender="Member Gender"
          ,birthyear= "05 - Member Details Member Birthday Year"
          )
df1 <-  mutate(df1, trip_id = as.character(trip_id)
                   ,bikeid = as.character(bikeid))
df2 <-  mutate(df2, trip_id = as.character(trip_id)
               ,bikeid = as.character(bikeid))
df3 <-  mutate(df3, trip_id = as.character(trip_id)
               ,bikeid = as.character(bikeid))
df4 <-  mutate(df4, trip_id = as.character(trip_id)
               ,bikeid = as.character(bikeid))
```

# Merging the dataset together
```{r echo=TRUE}
all_trips= bind_rows(df1,df2,df3,df4)
colnames(all_trips)
dim(all_trips)
summary(all_trips)
```

# Data cleaning
```{r echo=TRUE}
all_trips$date <- as.Date(all_trips$start_time)
all_trips$time <- format(as_datetime(all_trips$start_time),"%H")
all_trips$month <- format(as.Date(all_trips$start_time),"%b")
all_trips$day <- format(as.Date(all_trips$start_time),"%d")
all_trips$day_of_week <- format(as.Date(all_trips$start_time),"%A")
str(all_trips)
```

#selecting only relevant features
```{r echo=TRUE}
df <- select(all_trips,-c(start_time,end_time,bikeid,from_station_id,
                          to_station_id,birthyear))
```

# COMPARE

## trip duration with usertype
```{r echo=TRUE}
aggregate(df$tripduration~df$usertype,FUN = mean)
aggregate(df$tripduration~df$usertype,FUN = length)
```

## trip duration with gender
```{r echo=TRUE}
aggregate(df$tripduration~df$gender,FUN = mean)
aggregate(df$tripduration~df$gender,FUN = length)
```

## trip duration with day_week
```{r echo=TRUE}
#fixing the order
df$day_of_week <- ordered(df$day_of_week, levels=c("Sunday",
            "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
aggregate(df$tripduration~df$day_of_week,FUN = mean)
aggregate(df$tripduration~df$day_of_week,FUN = length)
```

## trip duration with month
```{r}
df$month <- ordered(df$month, levels=c("Jan","Feb", "Mar", "Apr", "May", "Jun",
                                       "Jul","Aug","Sep","Oct","Nov","Dec"))
aggregate(df$tripduration~df$month,FUN = mean)
aggregate(df$tripduration~df$month,FUN = length)
```

# ANALYSIS

## ridership data by usertype and weekday
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
graph_1 <- df %>%
  group_by(usertype,day_of_week) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(tripduration)) %>% 
  arrange(usertype,day_of_week)
```

#### Number of trips by usertype and weekday
```{r Number of trips by usertype and weekday, echo=TRUE}
ggplot(data = graph_1, aes(x=day_of_week,y=number_of_rides,fill = usertype)) +
  geom_col(position = "dodge")+labs(title = "Number of trips by usertype and weekday")
```

#### Mean trip duration by usertype and weekday
```{r Mean trip duration by usertype and weekday}
ggplot(data = graph_1, aes(x=day_of_week,y=avg_duration,fill = usertype)) +
  geom_col(position = "dodge")+labs(title = "Mean trip duration by usertype and weekday")
```

## ridership data by gender and weekday
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
graph_2 <- df %>%
  drop_na(gender) %>% 
  group_by(gender,day_of_week) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(tripduration)) %>% 
  arrange(gender,day_of_week)
```

#### Number of trips by gender and weekday
```{r Number of trips by gender and weekday}
ggplot(data = graph_2, aes(x=day_of_week,y=number_of_rides,fill = gender)) +
  geom_col(position = "dodge")+labs(title = "Number of trips by gender and weekday")
```

#### Mean trip duration by gender and weekday
```{r Mean trip duration by gender and weekday}
ggplot(data = graph_2, aes(x=day_of_week,y=avg_duration,fill = gender)) +
  geom_col(position = "dodge")+labs(title = "Mean trip duration by gender and weekday")
```

## ridership data by gender and usertype
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
graph_3 <- df %>%
  drop_na(gender) %>% 
  group_by(gender,usertype) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(tripduration)) %>% 
  arrange(gender,usertype)
```

#### Number of male and female users by usertype
```{r Number of male and female users by usertype}
ggplot(data = graph_3, aes(x=usertype,y=number_of_rides,fill = gender)) +
  geom_col(position = "dodge")+labs(title = "Number of male and female users by usertype")
```

## ridership data by week days and months
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
df$month <- ordered(df$month, levels=c("Jan","Feb", "Mar", "Apr", "May", "Jun",
                                       "Jul","Aug","Sep","Oct","Nov","Dec"))
df$day_of_week <- ordered(df$day_of_week, levels=c("Sunday", "Monday", "Tuesday",
                                  "Wednesday", "Thursday", "Friday", "Saturday"))
graph_4 <- df %>%
  group_by(day_of_week,month,usertype) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(tripduration)) %>% 
  arrange(day_of_week,month)
```

#### Number of subscribers by months and days of week
```{r Number of subscribers by months and days of week}
ggplot(data = graph_4, aes(x=day_of_week,y=number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")+
  facet_wrap(~month)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
  labs(title = "Number of subscribers by months and days of week")
```

## ridership data by week days and time
```{r echo=TRUE, message=FALSE, warning=FALSE, paged.print=FALSE}
df$day_of_week <- ordered(df$day_of_week, levels=c("Sunday",
                                                   "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
graph_5 <- df %>%
  group_by(day_of_week,time) %>% 
  summarise(number_of_rides = n(), avg_duration=mean(tripduration)) %>% 
  arrange(day_of_week,time)
```

#### Number of trips by time and days of week
```{r Number of trips by time and days of week, echo=TRUE}
ggplot(data = graph_5, aes(x=day_of_week,y=number_of_rides, fill = day_of_week)) +
  geom_col(position = "dodge")+
  facet_wrap(~time)+
  theme(axis.text.x = element_blank())+
  labs(title = "Number of trips by time and days of week")
```



# Observations

 * [From this plot][Number of trips by usertype and weekday] , we can see that number of rides for subscribers increase through working days and decreases on the weekends. While the customer number is high on weekends.
 
 * [From this plot][Mean trip duration by usertype and weekday] , we can see that ride duration is high for customers than subscribers. 

 * [From this plot][Number of trips by gender and weekday] , we can see that number of male riders increases through workdays and is at the lowest on weekends. Number of females riders are low compared to male riders and is fairly consistent throughout the week.
 
 * [Here][Mean trip duration by gender and weekday] , Average ride duration is higher for females compared to males and it is high on weekends and decreases through the working days. 
 
 * [This][Number of male and female users by usertype] plot shows that there are less number of female subscribers for the company.
 
 * [From this plot][Number of subscribers by months and days of week] , we can see that subscriptions start to increase from the month of April to October and is low on the months from November to March.
 
 * [Here][Number of trips by time and days of week] , we can see that the number of riders start to increase from morning 6 am, reaches a peak at 8 am, then decrease mid day and then again reaches peak at 4pm to 6pm, then decreases again.
 
 
 
# Conclusion

 * To increase the subscribers on weekends, weekend specific offers and discounts for subscribers taking bikes from/to entertainment centers or recreational activities should be introduced.

 * Provide special subscription offers to customers who have ride duration over a certain limit.

 * Advertisements and campaigns to increase the number of female users should be introduced.

 * Advertisements and discounts should be increased on the months from November to march to increase the number of users during this period.
